---
title: "physmerge: Manual"
author: "Pin"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

---

## Overview

`physmerge` collapses overlapping genomic signals into non-overlapping locus
blocks using a **forward sliding-window** approach — no LD reference panel
required.

### When to use this package

- You have GWAS summary statistics and want to identify independent loci
- You do not have access to a genotype-level LD reference panel for proper
  LD clumping
- You want a reproducible, fast, programmatic alternative to manual locus
  inspection

### Installation

```{r install, eval=FALSE}
devtools::install_github("Droideight/physmerge")
```

```{r library}
library(physmerge)
```

---

## Function reference

| Function | Input | Output | Purpose |
|---|---|---|---|
| `read_sumstat()` | File path | `list(data, reward)` | Read PLINK2 / GPCM / custom sumstats |
| `physical_merge()` | data frame | Blocks data frame | Run the merging algorithm |
| `annotate_blocks()` | Blocks + original data | Annotated blocks | Join full SNP info back to blocks |
| `export_snp_list()` | Annotated blocks | `.txt` or `.zip` | Export representative SNP IDs |

---

## Algorithm

The algorithm runs in three passes:

1. **Forward scan** — opens a block when a significant SNP is encountered.
   The block stays alive as long as the next SNP is within `window` bp.
   Whenever a more significant SNP is found inside the block, it becomes the
   new representative and the window resets.

2. **Collapse pass** — merges adjacent blocks whose representative SNPs
   (`rps_BP`) are fewer than `window` bp apart, keeping the more significant
   representative.

3. **Trim pass** — if any block's `end` still overlaps the next block's
   `start` after collapsing, `end` is trimmed to `start` of the next block,
   guaranteeing zero overlap.

---

## Case 1: Pre-processed data frame

Use `physical_merge()` directly when you already have a data frame with
`position` and `value` columns.

### Toy dataset

```{r toy_dataset}
set.seed(42)
n   <- 500
pos <- sort(sample(1:5e6, n))
pval <- runif(n, 0.01, 1)

# Locus 1: single clean peak around 1.05 Mb
peak1 <- which.min(abs(pos - 1050000))
pval[peak1 + c(-3,-2,-1,0,1,2,3)] <- c(1e-5, 1e-6, 1e-7, 1e-10, 1e-8, 1e-6, 1e-5)

# Locus 2: single clean peak around 2.5 Mb
peak2 <- which.min(abs(pos - 2500000))
pval[peak2 + c(-2,-1,0,1,2)] <- c(1e-6, 1e-7, 3e-9, 1e-7, 1e-6)

# Locus 3: two nearby sub-peaks around 4 Mb — should be collapsed into one block
peak3 <- which.min(abs(pos - 4000000))
pval[peak3 + c(-1,0,1)] <- c(1e-6, 2e-8, 1e-6)   # sub-peak A
pval[peak3 + c(3,4,5)]  <- c(1e-6, 5e-9, 1e-6)   # sub-peak B (closer → collapsed)

toy_df <- data.frame(position = pos, value = pval)
head(toy_df)
```

### Run merging

```{r merge_toy}
merged_toy <- physical_merge(
  toy_df,          # data frame with columns `position` and `value`
  sig_th = 5e-8,   # significance threshold
  window = 500000, # window size in base-pairs
  reward = "min"   # "min" for p-values; "max" for test statistics
)
merged_toy
```

We expect **3 blocks**: Locus 1, Locus 2, and Locus 3 (two sub-peaks
collapsed into one, with the stronger `5e-9` SNP as representative).

### Annotate blocks

`annotate_blocks()` joins the full original columns back to each block row.
For the toy dataset this adds nothing new, but for real data it brings in
`BETA`, `SE`, `REF`, `ALT`, etc.

```{r annotate_toy}
recovered_toy <- annotate_blocks(
  merged_toy,  # output from physical_merge()
  toy_df       # original data frame (must have `position` column)
)
recovered_toy
```

The `keep_*` parameters let you drop columns you do not need downstream:

```{r annotate_toy_slim, eval=FALSE}
# Drop start/end/rps_value — keep serial, rps_BP, rps_ID only
annotate_blocks(merged_toy, toy_df,
                keep_start     = FALSE,
                keep_end       = FALSE,
                keep_rps_value = FALSE)
```

---

## Case 2: PLINK2 summary statistics

`read_sumstat()` reads the file, filters to `TEST == "ADD"` rows (default
for PLINK2), and returns `list(data, reward)`.

```{r plink, eval=FALSE}
plink <- read_sumstat(
  "path/to/your.glm.linear",
  format      = "plink2",  # expects #CHROM, POS, ID, P
  test_filter = TRUE,       # keep only ADD rows (default for plink2)
  test_col    = "TEST",
  test_val    = "ADD",
  chrom       = NULL        # set e.g. chrom = 1:22 to restrict chromosomes
)

# plink is a list — data lives in plink$data
head(plink$data)

merged_plink    <- physical_merge(plink$data, 5e-8, 500000,
                                  reward = plink$reward)
recovered_plink <- annotate_blocks(merged_plink, plink$data)
head(recovered_plink)
```

Export the representative SNP list:

```{r plink_export, eval=FALSE}
# Single merged file
export_snp_list(recovered_plink, "snp_ids.txt")

# Verify
ids <- readLines("snp_ids.txt")
cat("Blocks:", nrow(recovered_plink), "| IDs written:", length(ids), "\n")
head(ids)
```

---

## Case 3: REML-GPCM per-SNP file

GPCM output has multiple p-value columns. Choose one with `value_col`.

```{r gpcm, eval=FALSE}
gpcm <- read_sumstat(
  "path/to/stage1_ch1_preclump.csv",
  format      = "gpcm",
  value_col   = "P_HPI",   # or "P_Direct", "P_Indirect", "P_TE"
  test_filter = FALSE       # GPCM files do not have a TEST column
)

head(gpcm$data)

merged_gpcm    <- physical_merge(gpcm$data, 5e-8, 500000,
                                 reward = gpcm$reward)
recovered_gpcm <- annotate_blocks(merged_gpcm, gpcm$data)
head(recovered_gpcm)

export_snp_list(recovered_gpcm, "snp_ids_gpcm.txt")

ids <- readLines("snp_ids_gpcm.txt")
cat("Blocks:", nrow(recovered_gpcm), "| IDs written:", length(ids), "\n")
head(ids)
```

---

## Case 4: Custom file format

When your file uses non-standard column names, use `format = "custom"` and
specify every column explicitly.

```{r custom, eval=FALSE}
custom <- read_sumstat(
  "path/to/results.txt",
  format    = "custom",
  chrom_col = "CHR",
  pos_col   = "BP",
  id_col    = "SNP",
  value_col = "P"
)
```

---

## Exporting by chromosome

When you need one SNP list per chromosome (e.g. for Swiss Army Knife RAW
extraction on UKB-RAP):

```{r export_bychr, eval=FALSE}
export_snp_list(recovered_plink, "snp_ids_by_chr.zip", by_chrom = TRUE)
# Produces: snp_ch1.txt, snp_ch2.txt, ... bundled in a zip
```

---

## Parameter reference

### `read_sumstat()`

| Parameter | Default | Description |
|---|---|---|
| `path` | — | Path to input file |
| `format` | `"plink2"` | `"plink2"` / `"gpcm"` / `"custom"` |
| `value_col` | format default | Column to use as merging value |
| `chrom_col` | format default | Chromosome column name |
| `pos_col` | format default | Position column name |
| `id_col` | format default | SNP ID column name |
| `test_filter` | format default | Filter rows by TEST column |
| `test_col` | `"TEST"` | TEST column name |
| `test_val` | `"ADD"` | Value to retain in TEST column |
| `chrom` | `NULL` | Chromosomes to keep, e.g. `1:22` |

### `physical_merge()`

| Parameter | Default | Description |
|---|---|---|
| `data` | — | data frame with `position` and `value` columns |
| `sig_th` | — | Significance threshold |
| `window` | — | Window size in base-pairs |
| `reward` | `"min"` | `"min"` for p-values; `"max"` for test statistics |

### `annotate_blocks()`

| Parameter | Default | Description |
|---|---|---|
| `blocks` | — | Output from `physical_merge()` |
| `data` | — | Original input data frame |
| `chrom_col` | `NULL` | Auto-detects `CHROM` / `#CHROM` |
| `id_col` | `NULL` | Auto-detects `ID` / `SNP` |
| `keep_serial` | `TRUE` | Include `serial` column |
| `keep_start` | `TRUE` | Include `start` column |
| `keep_end` | `TRUE` | Include `end` column |
| `keep_rps_BP` | `TRUE` | Include `rps_BP` column |
| `keep_rps_value` | `TRUE` | Include `rps_value` column |
| `keep_rps_ID` | `TRUE` | Include `rps_ID` column |

### `export_snp_list()`

| Parameter | Default | Description |
|---|---|---|
| `blocks` | — | Output from `annotate_blocks()` |
| `path` | — | Output file path (`.txt` or `.zip`) |
| `by_chrom` | `FALSE` | `TRUE` = per-chromosome zip |
| `id_col` | `NULL` | Auto-detects `rps_ID` then `rps_BP` |
